A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-12 tex/codatadp.tex
Repair definition of normal mu-term.

It remains to show the inverse of the previous lemma, that when a derivation tree with root node $\Ec$ contains a saturated node $\Fc$,
then $\Ec$ is satisfiable $\thD$.
First, we require additional terminology concerning $\mu$-terms.
If $u$ is a subterm of $\mu$-term $v$, then the \emph{interpretation of $u$ in $v$} is the $\mu$-term $v\interp{u}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \mu x. C( \vec t ) \in v$ indicates that $\mu x. C( \vec t )$ is the subterm of $v$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
v\interp{x}^M & = & v\interp{\mu x. C( \vec t )}^M & x \not\in M, x \mapsto \mu x. C( \vec t ) \in v  \\
v\interp{x}^M & = & x & \text{otherwise} \\
v\interp{\mu x. C( \vec t )}^M & = & \mu x. C( v\interp{ \vec t }^{M \cup \{ x \}} ) & x \not\in M \\
v\interp{\mu x. C( \vec t )}^M & = & x & \text{otherwise}
\end{array}\]
We will write $v\interp{u}$ as shorthand for $v\interp{u}^\emptyset$.
Given a $\mu$-term $v$ of the form $\mu x. C( \vec t )$, 
we say $u$ is a \emph{bisimilar subterm} of $v$ if and only if $v\interp{u} \vsim v$.
We say $v$ is \emph{normal} if and only if all of its bisimilar subterms are $x$,
and all of its strict subterms are also normal.
For instance, the $\mu$-term $v$ of the form $\mu x. \const{S}( \mu y. \const{S}( x ) )$ is not normal, 
since 
$v\interp{\mu y. \const{S}( x )}^\emptyset =$ 
$\mu y. \const{S}( v\interp{x}^{\{y\}} ) =$ 
$\mu y. \const{S}( v\interp{\mu x. \const{S}( \mu y. \const{S}( x ) ) }^{\{y\}} ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( v\interp{\mu y. \const{S}( x )}^{\{x,y\}} ) ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( y ) )$,
which is syntactically equivalent to $v$ for the renaming $\{ x \mapsto y, y \mapsto x \}$.
For any $\mu$-value $v$ of the form $\mu x. C( \vec w )$, we may recursively construct a normal value $\nf{v}$
(which we will call the \emph{normal form} of $v$)
that is a renaming of $v$ by replacing all of the bisimilar subvalues of $v$ with $x$,
and by replacing all of the non-normal subterms of $v$ with their corresponding normal forms.
For instance, $\nf{(\mu x. \const{S}( \mu y. \const{S}( x ) ))} = \mu x. \const{S}( x )$.

>>>>

It remains to show the inverse of the previous lemma, that when a derivation tree with root node $\Ec$ contains a saturated node $\Fc$,
then $\Ec$ is satisfiable $\thD$.
First, we require additional terminology concerning $\mu$-terms.
For a $\mu$-term $t_1$ with subterm $t_2$, 
the \emph{interpretation of $t_2$ in $t_1$} is the $\mu$-term $t_1\interp{t_2}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \mu x. C( \vec u ) \in t_1$ indicates that $\mu x. C( \vec u )$ is the subterm of $t_1$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
t_1\interp{x}^M & = & \mu x. C( t_1\interp{ \vec u}^{M \cup \{ x \}}) & x \not\in M, x \mapsto \mu x. C( \vec u ) \in t_1  \\
t_1\interp{x}^M & = & x & \text{otherwise} \\
t_1\interp{\mu x. C( \vec u )}^M & = & \mu x. C( t_1\interp{ \vec u }^{M \cup \{ x \}} ) & x \not\in M \\
t_1\interp{\mu x. C( \vec u )}^M & = & x & \text{otherwise}
\end{array}\]
We will often write $t_1\interp{t_2}$ as shorthand for $t_1\interp{t_2}^\emptyset$.
Given a $\mu$-term $t_1$ of the form $\mu x. C( u_1, \ldots, u_n )$ with subterm $t_2$,
we say $t_2$ is a \emph{bisimilar subterm} of $t_1$ if and only if it is of the form
$\mu y. C( v_1, \ldots, v_n )$ and $t_1\interp{ u_i } \sim t_1\interp{ v_i }$ for $i = 1,\ldots, n$.
We say $t_1$ is \emph{normal} if and only if it does not contain a strict bisimilar subterm
and all of its strict subterms are also normal.
For instance, the $\mu$-term $t_1$ of the form $\mu x. \const{S}( \mu y. \const{S}( y ) )$ is not normal, 
since $\mu y. \const{S}( y )$ is a bisimilar subterm of $t_1$.
In particular, note that their arguments have the same interpretation in $t_1$, 
that is,
$t_1\interp{\mu y. \const{S}( y )}^\emptyset =$ 
$\mu y. \const{S}( t_1\interp{y}^{\{y\}} ) =$ 
$\mu y. \const{S}( y )$ 
is syntactically equivalent to
$t_1\interp{y}^\emptyset =$ 
$\mu y. \const{S}( t_1\interp{y}^{\{y\}} ) =$ 
$\mu y. \const{S}( y )$.
The $\mu$-term $t_1$ of the form $\mu x. \const{S}( \mu y. \const{S}( x ) )$ is also not normal, 
since $\mu y. \const{S}( x )$ is a bisimilar subterm of $t_1$,
noting that
$t_1\interp{\mu y. \const{S}( x )}^\emptyset =$ 
$\mu y. \const{S}( t_1\interp{x}^{\{y\}} ) =$ 
$\mu y. \const{S}( t_1\interp{\mu x. \const{S}( \mu y. \const{S}( x ) ) }^{\{y\}} ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( t_1\interp{\mu y. \const{S}( x )}^{\{x,y\}} ) ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( y ) )$
is syntactically equivalent to $t_1\interp{x}^\emptyset = t_1$ for the renaming $\{ x \mapsto y, y \mapsto x \}$.

For any $\mu$-value $v$ of the form $\mu x. C( \vec w )$, we may recursively construct a normal value $\nf{v}$
(which we will call the \emph{normal form} of $v$)
that is a renaming of $v$ by replacing all of the bisimilar subvalues of $v$ with $x$,
and by replacing all of the non-normal subterms of $v$ with their corresponding normal forms.
For instance, $\nf{(\mu x. \const{S}( \mu y. \const{S}( x ) ))} = \mu x. \const{S}( x )$.
